Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-02 causal-bandit/icml-2015/general.tex
updates

We now consider the more general problem where the graph structure is known, but arbitrary.
The problem of learning the optimal intervention is made more complicated than the simple case because of 
the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\caption{}\label{fig:causalStructure_confounded}
\end{figure}

@@

We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ be the mixture distribution.
Let $a \in \calA$ be an intervention and define random variable
\eq{
Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
}
where $B_a \geq 0$ is some constant to be chosen subsequently and
\eq{
R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
}
Define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $\eta$, $T$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $I_t = a$ with probability $\eta$
\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE Choose $I = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
First note that $|Z_a(X_t)| \leq B_a$ and that
\eq{
\Var_Q[Z_a(X_t)] 
&\leq \EE_Q[Z_a(X_t)^2] \\
&\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] \\
&\leq m(\eta)\,.
}
Checking the expectation we have
\eq{
\EE_Q[Z] 
&= \EE_{\Ps_a}\left[Y \ind{R_a \leq B_a}\right] \\
&= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] \\
&= \mu_a - \beta_a\,,
}
where $0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a > B_a}$ is the negative bias. 
The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
\eq{
\beta_a \leq \Pn{a}{R_a > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
}
Let $\epsilon_a > 0$ be given by
\eq{
\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
}
Then by the union bound and Bernstein's inequality we have
\eq{
&\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \\ 
&\qquad\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
}
Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
\eq{
\mu_I \geq \hat \mu_I - \epsilon_I  
\geq \hat \mu_{a^*} - \epsilon_I  
\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
}
By the definition of the truncation
we have
\eq{
\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
}
and
\eq{
\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
}
Therefore for $C = \sqrt{2} + 4$ we have
\eq{
\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
}
as required.
\end{proof}

>>>>

We now consider the more general problem where the graph structure is known, but arbitrary.
The problem of learning the optimal intervention is made more complicated than the simple case because of 
the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual graph in Figure \ref{fig:causalStructure_confounded} where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\caption{}\label{fig:causalStructure_confounded}
\end{figure}

@@

We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
a truncated importance weighted estimator.

Let $a \in \calA$ be an intervention and define random variable
\eq{
Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
}
where $B_a \geq 0$ is some constant to be chosen subsequently and
\eq{
R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
}
Define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{|\calA|}$, $B \in [0,\infty)^{|\calA|}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $I_t = a$ with probability $\eta$
\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE Choose $I = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
First note that $|Z_{at}| \leq B_a$ and that
\eq{
\Var_Q[Z_{at}] 
&\leq \EE_Q[Z_{at}^2] \\
&\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] \\
&\leq m(\eta)\,.
}
Checking the expectation we have
\eq{
\EE_Q[Z_{at}] 
&= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] \\
&= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] \\
&= \mu_a - \beta_a\,,
}
where $0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}$ is the negative bias. 
The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
\eq{
\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
}
Let $\epsilon_a > 0$ be given by
\eq{
\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
}
Then by the union bound and Bernstein's inequality 
\eq{
&\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \\ 
&\qquad\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
}
Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
\eq{
\mu_I \geq \hat \mu_I - \epsilon_I  
\geq \hat \mu_{a^*} - \epsilon_I  
\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
}
By the definition of the truncation
we have
\eq{
\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
}
and
\eq{
\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
}
Therefore for $C = \sqrt{2} + 4$ we have
\eq{
\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
}
as required.
\end{proof}
